---
title: "Flights Delay Perfomence"
author: "Meghana Atluri"
format: 
  pdf:
    self-contained: true
---

# [DATA]{.underline}

<https://www.transtats.bts.gov/Fields.asp?gnoyr_VQ=FGJ>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  tidy = TRUE,
  comment = "#>",
  fig.align = 'center',
  echo = TRUE
)
```

```{r}
#Loding necessary libraries
library(dplyr)
library(readr)
library(janitor)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(rsample)
library(MASS)
library(pROC)
```

```{r}
#Importing file
flights_raw <- read_csv("/Users/meghanaatluri/Desktop/GIT/FLIGHTS/on_time_perf_dec2022.csv")
```

# [UNDERSTANDING OUR DATA]{.underline}

```{r}
#dimensions of dataset
dim(flights_raw)
```

```{r}
#Glimpse of data
glimpse(flights_raw)
```

```{r}
#column names
colnames(flights_raw)
```

```{r}
#datatype of col
str(flights_raw)
```

# [TIDY DATA]{.underline}

```{r}
#cleaning and renaming columns
flights_raw <- flights_raw |>
  clean_names() |>
  rename(
    Year = year,
    Quarter = quarter,
    Month = month,
    DayOfMonth = day_of_month,
    DayOfWeek = day_of_week,
    FlightDate = fl_date,
    OperatingAirline = op_unique_carrier,
    TailNumber = tail_num,
    FlightNumber = op_carrier_fl_num,
    OriginAirportCode = origin,
    OriginCity = origin_city_name,
    OriginState = origin_state_nm,
    DestAirportCode = dest,
    DestCity = dest_city_name,
    DestState = dest_state_nm,
    ScheduledDepTime = crs_dep_time,
    ActualDepTime = dep_time,
    DepDelayMinutes = dep_delay,
    ScheduledArrTime = crs_arr_time,
    ActualArrTime = arr_time,
    ArrDelayMinutes = arr_delay,
    Cancelled = cancelled,
    CancellationCode = cancellation_code,
    Diverted = diverted,
    CarrierDelayMinutes = carrier_delay,
    WeatherDelayMinutes = weather_delay,
    NASDelayMinutes = nas_delay,
    SecurityDelayMinutes = security_delay,
    LateAircraftDelayMinutes = late_aircraft_delay
  )
```

```{r}
#Drop TailNumber column
flights_cleaned = subset(flights_raw, select = -c(TailNumber) )
```

```{r}
#Checking number of NA values in each column
count_na <- function(x) {
  sum(is.na(x))
}
apply(flights_cleaned, MARGIN = 2, FUN = count_na)
```

```{r}
#Handling NA : Imputing 9999(flag) in delay min for cancelled flights
flights_cleaned <- flights_cleaned %>%
  mutate(
    DepDelayMinutes = if_else(is.na(DepDelayMinutes) & Cancelled == 1, 9999, DepDelayMinutes),
    ArrDelayMinutes = if_else(is.na(ArrDelayMinutes) & Cancelled == 1, 9999, ArrDelayMinutes)
    )
```

```{r}
#New Columns 
flights_cleaned <- flights_cleaned %>%
  mutate(
    CancelledFlag = if_else(Cancelled == 1, 1, 0),
    DivertedFlag = ifelse(Diverted == 1, 1, 0)
  )
```

```{r}
#Handling NA:Replace NA with "Not Canceled" (when Cancelled == 0).
flights_cleaned <- flights_cleaned %>%
  mutate(
    CancellationCode = if_else(is.na(CancellationCode) & Cancelled == 0, "Not Canceled", as.character(CancellationCode))
  )

```

```{r}
#Handling NA

flights_cleaned <- flights_cleaned %>%
  mutate(
    # If ArrivalDelay is 0 or negative (on-time or early), set all delay columns to 0
    CarrierDelayMinutes = if_else(ArrDelayMinutes <= 0 & Cancelled == 0 & Diverted == 0, 0, CarrierDelayMinutes),
    WeatherDelayMinutes = if_else(ArrDelayMinutes <= 0 & Cancelled == 0 & Diverted == 0, 0, WeatherDelayMinutes),
    NASDelayMinutes = if_else(ArrDelayMinutes <= 0 & Cancelled == 0 & Diverted == 0, 0, NASDelayMinutes),
    SecurityDelayMinutes = if_else(ArrDelayMinutes <= 0 & Cancelled == 0 & Diverted == 0, 0, SecurityDelayMinutes),
    LateAircraftDelayMinutes = if_else(ArrDelayMinutes <= 0 & Cancelled == 0 & Diverted == 0, 0, LateAircraftDelayMinutes),
    
    # For cancelled flights, assign 9999(flag) for delay columns (since flight didn't occur)
    CarrierDelayMinutes = if_else(is.na(CarrierDelayMinutes) & Cancelled == 1, 9999, CarrierDelayMinutes),
    WeatherDelayMinutes = if_else(is.na(WeatherDelayMinutes) & Cancelled == 1, 9999, WeatherDelayMinutes),
    NASDelayMinutes = if_else(is.na(NASDelayMinutes) & Cancelled == 1, 9999, NASDelayMinutes),
    SecurityDelayMinutes = if_else(is.na(SecurityDelayMinutes) & Cancelled == 1, 9999, SecurityDelayMinutes),
    LateAircraftDelayMinutes = if_else(is.na(LateAircraftDelayMinutes) & Cancelled == 1, 9999, LateAircraftDelayMinutes)
  )

```

```{r}
# Imputing median for delay columns where CancelledFlag is not TRUE
flights_cleaned <- flights_cleaned %>%
  mutate(
    CarrierDelayMinutes = ifelse(CancelledFlag != TRUE & is.na(CarrierDelayMinutes),
                                 round(median(CarrierDelayMinutes[CancelledFlag != TRUE], 
                                              na.rm = TRUE)),
                                 CarrierDelayMinutes),
    WeatherDelayMinutes = ifelse(CancelledFlag != TRUE & is.na(WeatherDelayMinutes),
                                 round(median(WeatherDelayMinutes[CancelledFlag != TRUE], 
                                              na.rm = TRUE)),
                                 WeatherDelayMinutes),
    NASDelayMinutes = ifelse(CancelledFlag != TRUE & is.na(NASDelayMinutes),
                             round(median(NASDelayMinutes[CancelledFlag != TRUE], na.rm = TRUE)),
                             NASDelayMinutes),
    SecurityDelayMinutes = ifelse(CancelledFlag != TRUE & is.na(SecurityDelayMinutes),
                                  round(median(SecurityDelayMinutes[CancelledFlag != TRUE],
                                               na.rm = TRUE)),
                                  SecurityDelayMinutes),
    LateAircraftDelayMinutes = ifelse(CancelledFlag != TRUE & is.na(LateAircraftDelayMinutes),
                                      round(median(LateAircraftDelayMinutes[CancelledFlag != TRUE], na.rm = TRUE)),
                                      LateAircraftDelayMinutes)
  )

```

```{r}
#Checking number of NA values in each column
count_na <- function(x) {
  sum(is.na(x))
}
apply(flights_cleaned, MARGIN = 2, FUN = count_na)
```

```{r}
# Convert columns to factors
flights_cleaned <- flights_cleaned %>%
  mutate(
    Year = as.factor(Year),
    Quarter = as.factor(Quarter),
    Month = as.factor(Month),
    DayOfMonth = as.factor(DayOfMonth),
    DayOfWeek = as.factor(DayOfWeek),
    OperatingAirline = as.factor(OperatingAirline),
    FlightNumber = as.factor(FlightNumber),
    OriginAirportCode = as.factor(OriginAirportCode),
    DestAirportCode = as.factor(DestAirportCode),
    CancellationCode = as.factor(CancellationCode),
    OriginCity = as.factor(OriginCity),
    OriginState = as.factor(OriginState),
    DestCity = as.factor(DestCity),
    DestState = as.factor(DestState)
  )
```

```{r}
# Convert Date columns to Date 
flights_cleaned <- flights_cleaned %>%
  mutate(
    FlightDate = format(mdy_hms(FlightDate), "%Y-%m-%d")
  )

```

```{r}
# Convert Date columns to Date
flights_cleaned$ScheduledDepTime <- format(as.POSIXct(sprintf("%04s", as.character(flights_cleaned$ScheduledDepTime)), format = "%H%M", tz = "UTC"), "%H:%M")
flights_cleaned$ActualDepTime <- format(as.POSIXct(sprintf("%04s", as.character(flights_cleaned$ActualDepTime)), format = "%H%M", tz = "UTC"), "%H:%M")

flights_cleaned$ScheduledArrTime <- format(as.POSIXct(sprintf("%04s", as.character(flights_cleaned$ScheduledArrTime)), format = "%H%M", tz = "UTC"), "%H:%M")
flights_cleaned$ActualArrTime <- format(as.POSIXct(sprintf("%04s", as.character(flights_cleaned$ActualArrTime)), format = "%H%M", tz = "UTC"), "%H:%M")
```

```{r}
# Convert to Numeric
flights_cleaned$DepDelayMinutes <- as.numeric(flights_cleaned$DepDelayMinutes)
flights_cleaned$ArrDelayMinutes <- as.numeric(flights_cleaned$ArrDelayMinutes)
```

```{r}
# Convert to Factors with levels
flights_cleaned <- flights_cleaned %>%
  mutate(
    Cancelled = factor(Cancelled, levels = c(0, 1)),
    Diverted = factor(Diverted, levels = c(0, 1)),
    CancelledFlag = factor(CancelledFlag , levels = c(0, 1)),
    DivertedFlag = factor(DivertedFlag, levels = c(0, 1) )
  )
```

```{r}
# Convert delay columns to Numeric
flights_cleaned <- flights_cleaned %>%
  mutate(
    CarrierDelayMinutes = as.numeric(CarrierDelayMinutes),
    WeatherDelayMinutes = as.numeric(WeatherDelayMinutes),
    NASDelayMinutes = as.numeric(NASDelayMinutes),
    SecurityDelayMinutes = as.numeric(SecurityDelayMinutes),
    LateAircraftDelayMinutes = as.numeric(LateAircraftDelayMinutes)
  )
```

```{r}
#Ckecking datatype
str(flights_cleaned)
```

# [Exploratory Data Analysis (ED]{.underline}A)

```{r}
# Create dataset for Cancelled flights
cancelled_flights <- flights_cleaned %>%
  filter(Cancelled == 1) 
```

```{r}
#Create dataset for all flights(- Cancelled flights)
flights_delay <- flights_cleaned %>%
  filter(Cancelled == 0) 
```

```{r}
#Summary
summary(flights_delay)
```

#### **1.Understanding Flight Perfomence**

```{r}
# Total number of flights
total_flights <- nrow(flights_cleaned)

# Number of cancelled flights
cancelled_flights <- sum(flights_cleaned$Cancelled == 1)

# Number of diverted flights
diverted_flights <- sum(flights_cleaned$Diverted == 1)

# Percentages
cancelled_pct <- round((cancelled_flights / total_flights) * 100, 2)
diverted_pct <- round((diverted_flights / total_flights) * 100, 2)

# Summary table
summary_df <- data.frame(
  Metric = c("Total Flights", "Cancelled Flights", "Diverted Flights"),
  Count = c(total_flights, cancelled_flights, diverted_flights),
  Percent = c(100, cancelled_pct, diverted_pct)
)
summary_df

```

```{r}
# Bar chart for counts (corrected)
ggplot(summary_df, aes(x = Metric, y = Count, fill = Metric)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count) )+
  labs(title = "Flight Summary: Total, Cancelled, Diverted",
       y = "Number of Flights",
       x = "") +
  theme_minimal() +
  theme(legend.position = "none")

```

There were 557095 flights in Dec 2022. About 5.49% were cancelled and 0.23% were diverted. The majority of flights (94.25%) were completed as scheduled.

#### **2.Analyze delay patterns by airline**

```{r}
#Avd Departure Delay (in min) by airline 
avg_dep_delay <- flights_cleaned %>%
  filter(Cancelled == 0) %>%
  group_by(OperatingAirline) %>%
  summarise(AvgDepDelay = round(mean(DepDelayMinutes, na.rm = TRUE), 2)) %>%
  arrange(desc(AvgDepDelay))
avg_dep_delay
```

```{r}
# Bar plot
ggplot(avg_dep_delay, aes(x = reorder(OperatingAirline, -AvgDepDelay), y = AvgDepDelay)) +
  geom_col(fill = 'steelblue') +
  labs(title = "Average Departure Delay by Airline", x = "Airline", y = "Avg Departure Delay (in mins)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Airlines F9 has the highest average departure delays (\~31 mins), while OH , YX had the lowest (\~5 mins) indicating relative airline punctuality.

#### **3.Airports with frequent delays**

```{r}
# Top 10 OriginAirports by Avg Arrival Delay
top_origin_delay <- flights_cleaned %>%
  filter(Cancelled == 0) %>%
  group_by(OriginAirportCode) %>%
  summarise(AvgArrDelay = round(mean(ArrDelayMinutes, na.rm = TRUE), 2)) %>%
  arrange(desc(AvgArrDelay)) %>%
  head(10)
top_origin_delay
```

```{r}
# Bar chart
ggplot(top_origin_delay, aes(x = reorder(OriginAirportCode, -AvgArrDelay), y = AvgArrDelay)) +
  geom_col(fill = 'pink') +
  labs(title = "Top 10 Origin Airports by Avg Arrival Delay", x = "Origin Airport", y = "Avg Arrival Delay (mins)")
```

Airports BIH (Bishop, CA), SCK (Stockton,CA) show higher average arrival delays (\~66+ mins), possibly indicating carrier or weather issues at these locations.

#### 4.Impact of day of week on delays

```{r}
#Arrival Delay by dayofweek
del_day_week <- flights_cleaned %>%
  filter(Cancelled == 0) %>%
  group_by(DayOfWeek) %>%
  summarise(AvgArrDelay = round(mean(ArrDelayMinutes, na.rm = TRUE), 2)) %>%
  arrange(desc(AvgArrDelay)) 
del_day_week
```

```{r}
#Horizontal Bar Graph
ggplot(del_day_week, aes(x = reorder(factor(DayOfWeek), AvgArrDelay), y = AvgArrDelay)) +
  geom_col(fill = "purple", alpha = 0.6) +
  labs(title = "Arrival Delay by Day of the Week",
       x = "Day of Week (1=Mon, 7=Sun)",
       y = "Average Arrival Delay (mins)") +
  theme_minimal() +
  coord_flip()
```

Delays appear slightly higher on weekends (Fri - Sun) likely due to heavier passenger loads and weekend travel surges.

#### 5.Reasons for cancellations

```{r}
#Cancellation reason
cancel_reasons <- flights_cleaned %>%
  filter(Cancelled == 1) %>%
  group_by(CancellationCode) %>%
  summarise(Count = n()) %>%
  mutate(Percent = round((Count / sum(Count)) * 100, 2))
cancel_reasons
```

```{r}
# Bar plot
ggplot(cancel_reasons, aes(x = CancellationCode, y = Count, fill = CancellationCode)) +
  geom_col() +
  labs(title = "Cancellation Reasons Distribution", x = "Cancellation Code", y = "Count") +
  theme(legend.position = "none")
```

Most flight cancellations (61%) were due to Carrier-related reasons (Code A), followed by Weather (38%). NAS-related issues (Code C) accounted for only 1.6%.

#### 6.Relationship between departure and arrival delays

```{r}
# Fit the linear regression model
lm_model <- lm(ArrDelayMinutes ~ DepDelayMinutes, data = flights_cleaned %>% filter(Cancelled == 0))

# Summary of the model
lm_summary <- summary(lm_model)
lm_summary
```

```{r}
#Regression line plot
ggplot(flights_cleaned %>% filter(Cancelled == 0), aes(DepDelayMinutes,ArrDelayMinutes )) +
  geom_point(alpha=0.7, size=0.9) +
  geom_smooth(method="lm", se=FALSE, size=1)
```

-   R² = 0.9474 → Very strong linear relationship (94.74% of the variation in Arrival Delay based on Departure Delay. )

-   Residual Std. Error = 14.94 min (predictions deviate \~15 mins from actual arrival delay )

-   Highly significant variables, both intercept and departure delay are statistically relevant in predicting arrival delay.

#### 7.Impact of weather and other factors on delays

```{r}
delay_factors <- flights_cleaned %>%
  filter(Cancelled == 0, Diverted == 0) %>%
  summarise(
    Carrier = sum(CarrierDelayMinutes, na.rm = TRUE),
    Weather = sum(WeatherDelayMinutes, na.rm = TRUE),
    NAS = sum(NASDelayMinutes, na.rm = TRUE),
    Security = sum(SecurityDelayMinutes, na.rm = TRUE),
    LateAircraft = sum(LateAircraftDelayMinutes, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Cause", values_to = "TotalMinutes")

delay_factors
```

```{r}
# Bar plot
ggplot(delay_factors, aes(x = reorder(Cause, -TotalMinutes), y = TotalMinutes)) +
  geom_col(fill = 'darkgreen') +
  labs(title = "Total Delay Minutes by Cause", x = "Cause of Delay", y = "Total Delay Minutes")

```

Late aircraft connections and NAS (Airspace system) issues contribute the largest share of total delays, while Security causes very minimal delays.

# [MODELING]{.underline}

#### Classification Model — Predict Delay

```{r}
# DepDelayFlag (0 = No delay, 1 = Delay)
flights_delay$DepDelayFlag <- ifelse(flights_delay$DepDelayMinutes > 15, 1, 0)
flights_delay$DepDelayFlag <- factor(flights_delay$DepDelayFlag, levels = c(0, 1))

```

Created a new column called DepDelayFlag if delay of more than 15 minutes it assigns a value of 1 (delayed flight) else 0. Its a factor with two levels: 0 and 1

```{r}
# Split data
set.seed(123)
split <- initial_split(flights_delay, prop = 0.8, strata = DepDelayFlag)
train_data <- training(split)
test_data  <- testing(split)
```

Splits the dataset into two parts: a training and test set where 80% of the data is training set, and the remaining 20% will be used for testing.

```{r}
# Fit a GLM model 
glm_model <- glm(DepDelayFlag ~  CarrierDelayMinutes + WeatherDelayMinutes + NASDelayMinutes + SecurityDelayMinutes + LateAircraftDelayMinutes  + DayOfWeek  + OperatingAirline, 
                 family = binomial, 
                 data = train_data)
# Summarize the model
summary(glm_model)
```

-   Logistic regression to predict whether there is a departure delay using several predictor variables.

-   Dependent variable - DepDelayFlag 0 indicating no delay and 1 indicating a delay.

-   Predictor variables: CarrierDelayMinutes ,WeatherDelayMinutes, NASDelayMinutes, SecurityDelayMinutes, LateAircraftDelayMinutes, DayOfWeek, OperatingAirline.

-   The intercept is -3.035, meaning when all predictors are zero, the log-odds of a delay are negative (the probability of a delay is low).

-   The positive coefficient (CarrierDelayMinutes ,WeatherDelayMinutes, NASDelayMinutes, SecurityDelayMinutes, LateAircraftDelayMinutes) indicates that for every additional minute of carrier delay, the log-odds of a delay increase. This predictor is highly significant.

-   Predictors (CarrierDelayMinutes ,WeatherDelayMinutes, NASDelayMinutes, SecurityDelayMinutes, LateAircraftDelayMinutes )have a very strong impact on the likelihood of a delay, as indicated by their very small p-values.

-   Day of the week and the operating airline are also significant in many cases but not for all values (e.g DayOfweek2 , DayOfWeek5 are not significant).

```{r}
step_model <- step(glm_model, direction = "backward")
summary(step_model)
```

-   Did backward stepwise selection on GLM model using step().

-   Backward stepwise selection retained all predictors in the model, meaning each variable contributed to improving model fit (AIC). Removing any variable increased AIC, suggesting its relevance to predicting flight delays.

-   Although some predictors were not individually significant (p \> 0.05), backward stepwise selection retained them because their inclusion improved model fit as measured by AIC.

```{r}
# Predict on test data
pred_probs <- predict(step_model, newdata = test_data, type = "response")
pred_class <- ifelse(pred_probs > 0.4, 1, 0)
```

Predicts the probability that flight is delayed or each row in test_data. If predicted probability \> 0.5 its Delayed If predicted probability ≤ 0.5 its not delayed.

```{r}
# Confusion Matrix
confusion_matrix <- table(Predicted = pred_class, Actual = test_data$DepDelayFlag)
confusion_matrix
```

TN - 76587 ; FP - 1276 ; FN - 5649 ; TP - 21791

```{r}
# Calculate accuracy
accuracy <- mean(pred_class == test_data$DepDelayFlag)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))
```

The model correctly predicted 93.42% of the flight delays.

```{r}
# Precision, Recall, F1-Score Calculation
precision <- confusion_matrix[2, 2] / sum(confusion_matrix[2, ])
recall <- confusion_matrix[2, 2] / sum(confusion_matrix[, 2])
f1_score <- 2 * (precision * recall) / (precision + recall)

print(paste("Precision: ", round(precision, 4)))
print(paste("Recall: ", round(recall, 4)))
print(paste("F1-Score: ", round(f1_score, 4)))
```

-   A precision of 0.9447 means that about 94.47% of the flights predicted to be delayed were indeed delayed.

-   A recall of 0.7941 indicates that the model correctly identified about 79.41% of the flights that were actually delayed.

-   An F1-Score of 0.8629 shows that the model has a good balance between precision and recall.

```{r}
pred_probs <- predict(step_model, newdata = test_data, type = "response")
pr_curve <- roc(test_data$DepDelayFlag, pred_probs, plot = TRUE, print.auc = TRUE, col = "blue")
```

```{r}

# Get predicted probabilities from the model
pred_probs <- predict(step_model, newdata = test_data, type = "response")

# Create ROC curve object
roc_curve <- roc(test_data$DepDelayFlag, pred_probs)

# Calculate AUC
roc_auc <- auc(roc_curve)

# Print AUC
print(paste("ROC AUC: ", round(roc_auc, 4)))

```

```{r}
library(readr)
write_csv(flights_cleaned, "/Users/meghanaatluri/Desktop/GIT/FLIGHTS/
          flights_cleaned.csv")
write_csv(flights_delay, "/Users/meghanaatluri/Desktop/GIT/FLIGHTS/
          flights_delay.csv")
```

# EXPORTING FROM R

```{r}
coefs <- summary(step_model)$coefficients
coefs_df <- data.frame(Feature = rownames(coefs), Coefficient = coefs[, "Estimate"])
write.csv(coefs_df, "/Users/meghanaatluri/Desktop/GIT/FLIGHTS/feature_importance.csv", row.names = FALSE)

```

```{r}
pred_df <- data.frame(
  ActualDelayFlag = test_data$DepDelayFlag,
  PredictedProb = pred_probs,
  PredictedClass = pred_class,
  OperatingAirline = test_data$OperatingAirline,
  OriginAirportCode = test_data$OriginAirportCode
)

write.csv(pred_df, "/Users/meghanaatluri/Desktop/GIT/FLIGHTS/predictions.csv", row.names = FALSE)


